---
title: "Homeostatic Disregulation Example"
author: "Bryce Bartlett"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\Homeostatic Disregulation Example{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message=FALSE)
```

## Dataset identification and subsetting.

In this vignette, I use a previously subset of the NHANES with particular biomarkers, which includes individuals 30 to 75 years old, and excludes pregnant women.

```{r}
library(dplyr)
library(bioage) #beta version currently -- topic of vignette
```

```{r, echo=FALSE}
#This codeblock loads data from my local machine that has previously been downloaded and cleane. The Original data is not available.
library(haven)
nhanes = read_dta('C:/Users/bjb40/Box Sync/Bartlett_BioAge/Data/nhanes/Stata/NHANESforBA.dta')

##limit by excluding pregnant woman and constraining the ages
nhanes = nhanes %>%
  filter(pregnant==0) %>%
  mutate(seqn=ifelse(samp=='nhanes3',paste0('33333',seqn),seqn)) #liminate duplicate sequence numbers

```

## Developing training fit for homeostatic disregulation.

For homeostatic disregulation, the constructed varialbe is based on a malhanobis distance statistic, which is theoretically the distance between observations and a hypothetically healthy, young cohort.

The "filter" argument is how the "healthy" restrictions are input. In this example, we train separately for men and women who are between the ages of 20 and 30, and have observe biomarker data within clinically accpetable distributions.

The table below identifies the sex-specific restrictions base on clinical guidelines, and the variables in the nhanes dataset. For clinical guidelines, we relied upon the ranges reported by the Mayo Clinic website (http://www.mayomedicallaboratories.com/test-catalog/Clinical+and+Interpretive/8340)[http://www.mayomedicallaboratories.com/test-catalog/Clinical+and+Interpretive/8340] (last vistited May 28, 2018).

|Variable|Description|Female Healthy Range| Male Healthy Range|
|:-------|:----------|:---------------------------|:-----------|
|seqn|Unique individual identifier||
|sex|gender (1=female; 2=male)||
|**Biomarkers**||
|albumin|Albumin (g/DL)|3.5-5,3.5-5|
|lnalp|Alkaline Phosphate (log U/L)|log(37)-log(98)|log(45)-log(115)|
|bun|Blood Urea Nitrogen|6-21|8-24|
|cmv|CMV|<2|<2|  
|lncreat|Creatinine (log)||
|lncrp|CRP (log)||
|hba1c|Glycalated Hemoglobin (%)||
|lymphpct|Lymphocite Percent|limited to ages 45-75|
|lnwbc|White Blood Cell Count (log)||
|lnua|Uric Acid (log)||
|sbp|Systolic Blood Pressure||
|mcv|Mean Corpuscular Volume||
|totchol|Total Cholesterol|limited to ages 30-55|


```{r}
biomarkers = c('albumin','lnalp','bun','crp')
filter.female = list(
              albumin='albumin>=3.5 & albumin<=5',
              lnalp = 'exp(lnalp)>=37 & exp(lnalp)<=98',
              bun = 'bun>=6 & bun<=21',
              crp = 'crp<2'
              )

hdtrain.female = hd(nhanes %>% filter(age>=18 & age<=35 & sex ==1),
            biomarkers=biomarkers,
            filter=filter.female)


```

Having calculated the training data, we can project homeostatic disregulation into out-of-sample data (i.e. the full dataset).

```{r}
hdvalidate.female = hd(nhanes %>% filter(sex==1),
                       biomarkers=biomarkers,
                       fit=hdtrain.female$fit)


plot(hdvalidate.female$data[,c('age','hd')])
cor(hdvalidate.female$data[,c('age','hd')],use='pairwise.complete.obs')

```


