---
title: "Biological Age Training Example"
author: "Bryce Bartlett"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\Biological Age Training Example{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette (1) uses the NHANES package to download publicly curated data with biomarkers; (2) splits the data in half for a training and validation set. Once the data is prepared, it uses the bioage package to (1) train biological age on the training subset; and (2) uses the estimated fit to calculate biological age for the test data.

## Gathering Data

This vignette uses the package nhanesA (not developed or associated with the bioage package) to draw publicly available data collected by the US Governemnt's National Health and Nutrition Survey (NHANES).

As noted in the code comments, this vignette uses the unassociated (but excellent) library, "nhanesA". 

```{r, warning=FALSE,message=FALSE}
#load library to pull nhanes data
#nhanesA is a nonaffiliated library, but is connected to a 
#commonly used repository that contains
#biomarkers
library(nhanesA) 

#draw standard biochemestry profile from NHANES 2001-2002 
#the excelelnet nhanesA library has information on how to 
#select and identify variables
nhdata = nhanes('L40_B')

#merge this with demographic data, which includes age
#SEQN is the individual identifier for nhanes data
nhdata = merge(nhdata,nhanes('DEMO_B'),by='SEQN')

#Subset the data
#here is a peak at the data, including 
#the sequence number, select biomarkers, and gender
print(head(nhdata[,c('SEQN',#sequence number
                     'LBXSAL', #Albumin
                     'RIAGENDR', #sex
                     'RIDAGEYR' #age (in years)
                     )]))

```


## Splitting the Data.

In this section I split the 2002 NHANES data to illustrate both how to train data, and how to use estimated parameters to generate out-of-sample biological age values.

```{r, results='hide',warning=FALSE,message=FALSE}
#select the first 1/2 of the data
train = nhdata[1:ceiling(nrow(nhdata)/2),]
#put the rest of the data in validation dataset
validate = nhdata[!nhdata$SEQN %in% train$SEQN,]
```

## Using the bioage package to train the data.

To train and validate the data, I use 5 biomarkers as oultined below:

|Variable Name|Description|
|:------------|:----------|
|LBXSAL|Albumin (g/dL)|
|LBDSAPSI|Alkaline Phosphate (U/L)|
|LBXSBU|Blood Urea Nitrogen (mg/dL)|
|LBXSCH|Cholesterol (mg/dL)|
|LBDSCR|Creatnine (mg/dL)|

**Note:** While this example includes individuals of all ages and pregnant women, this approach will likely give poor results for biological age. Researchers should therefore exclude these individuals in their analysis.

```{r, results='hide',warning=FALSE,message=FALSE}

library(bioage)

kdmtrain = kdm_calc(data=train,
                    agevar='RIDAGEYR',
                    biomarkers=c('LBXSAL','LBDSAPSI','LBXSBU',
                                 'LBXSCH','LBDSCR')
  
)
 

```

With the trained data, we can export the training parameters as a dataset, or extract the trained dataset.

The biological age variable is named as the concatenation of agevar with the prefix of 'bio'. 

Since the agevar is "RIDAGEYR", the calculated biological age is 'bioRIDAGEYR'. Here is a peak at the data with the biological age variable:

```{r}
newdata = extract_data(kdmtrain)

print(head(newdata[,c('SEQN',#sequence number
                     'LBXSAL', #Albumin
                     'RIAGENDR', #sex
                     'RIDAGEYR', #age (in years)
                     'bioRIDAGEYR' #biological age
                     )]))
```

We can also extract and print the parameters for the KDM algorithm. These parameters have special meanings relative to the algorithm as explained in papers discussing biological aging.

```{r}
params = extract_fit(kdmtrain)
print(params)
```

## Using trained data to estimate out-of-sample biological ages.

Finall, we can use the trained parameters to estimate biological ages for out of sample data. For this, the kdm_calc fucntion needs to be given training data, as follows:

```{r}

kdmvalidate = kdm_calc(data=validate,
                    agevar='RIDAGEYR',
                    biomarkers=c('LBXSAL','LBDSAPSI','LBXSBU',
                                 'LBXSCH','LBDSCR'),
                    fit=kdmtrain$fit #this argument supplies the fit
                    )

#the same functions apply as to the trianing data
new_validate = extract_data(kdmvalidate)

print(head(new_validate[,c('SEQN',#sequence number
                     'LBXSAL', #Albumin
                     'RIAGENDR', #sex
                     'RIDAGEYR', #age (in years)
                     'bioRIDAGEYR' #biological age
                     )]))



```

