---
title: "Biological Age Training Example"
author: "Bryce Bartlett"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\Biological Age Training Example{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette (1) uses the NHANES package to download publicly curated data with biomarkers; (2) splits the data in half for a training and validation set. Once the data is prepared, it uses the bioage package to (1) train biological age on the training subset; and (2) uses the estimated fit to calculate biological age for the test data.

## Gathering Data

This vignette uses the package nhanesA (not developed or associated with the bioage package) to draw publicly available data collected by the US Governemnt's National Health and Nutrition Survey. 

```{r}
#load library
library(nhanesA)

#draw demographic information 


#draw labs


#perform


```


## Splitting the Data.

```{r}

```

## Using the bioage package to train the data.

```{r}

```

## Using trained data to estimate out-of-sample biological ages.

```{r}

```

